Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-18 paper.tex
Rough pass on Discussion

Our final example concerns [TODO motivate this example; why is 
this interesting/useful?]
% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, with a known time to recombination,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the lefthand or
righthand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.
However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.
This approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution and context to a dataset of genetic variation.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently.

@@

% Although we can associate a likelihood with any valid tree sequences,
% there are some caveats. By treating each edge independently we can
% naturally deal with the various degrees of completeness to which inheritance
% is documented by the different ARG inference methods: e.g., polytomies, absence
% of locally unary coalescence nodes, unpreserved identify of internal nodes
% across marginal trees, etcetera.
% However, many of these imprecisions violate the implicit
% assumptions of the likelihood computation.
% \comment{Do we want these remarks in here, or as part of the Discussion?}
% 
% For example, the described algorithm assumes the
% presence of ``locally unary'' nodes to correctly identify the
% full extent of all lineages going back through time and therefore recombination events.
% Without unary nodes though, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. This can be mitigated by limiting
% the number of 'inferrable' recombination events per coalescence event to 1
% (see fig \ref{sup:fig:rec-correction}).
% \comment{I'm not editing this in since I'm not sure I've got it right:
% I think that as written, our algorithm would count an extra recombination
% and and extra coalescence for any edge that got split up by removing the unary bits.
% This could be corrected for in the algorithm, but it would require extra bookkeeping,
% and make the likelihood not so easily a product over edges. (?)}
% 
% \comment{this was previously part of the discussion, have grouped it here.}
% Although their presence is crucial to our ability to embed any local tree in the
% ARG, the concept and their importance are still relatively new (but see
% \citet{wong_general_2023}).
% We have briefly touched upon the observation that in
% the presence of unary nodes we seem to augment the left-to-right logic of the SMC
% with some additional information. Yet, we believe that the topic of locally
% unary nodes in general requires further exploration.
% And more specifically, we see it as future work to fully quantify to what
% extent the absence of these locally unary nodes will affect the shape of the
% likelihood surface given any set of parameters.

@@

The key contribution of this paper is to rederive the classical KYF
ARG likelihood function in a way that supports partial and imprecise
input. 
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, acknowledging that we 
do not necessarily observe or estimate these events, we calculate 
the possible contributions of various type of event to the inheritance
relationships that are provided as input.

[JK: this needs significant work]
Here, we have introduced a scalable algorithm to compute likelihoods
under the SMC for a general class of ARGs. By formulating the SMC backwards in time
we were able to associate a single likelihood value with each edge in the ARG.
The compatibility of the algorithm with the \texttt{tskit} library allows us to
exploit its associated efficiencies both for the current work and its future
applications. This work will contribute significantly
to ARG-based analysis becoming a standard part of the population
and statistical genetics toolkit.
In particular, we see two main applications for the work presented here.
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
model and an associated set of parameters, will help bridge the gap between the tremendous recent
advances in ARG inference and the current state of the art statistical inference of
past demography.
The second and probably the most important application of the ability to associate a more
general class of ARGs to a principled model is the ability to boost the current efforts
in ARG inference. 
% great starting point for MCMC
Any inferred ARG can be used a good starting point
for a future MCMC-sampler. This potentially implies a massive runtime reduction by
reducing the necessary burn-in period. Furthermore, by integrating out the exact time
to the recombination event we further restrict the size of the ARG space that needs to
be explored. This MCMC-sampler would again rely on the tree sequence data structure to
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
% hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
To further improve scalability, hybrid ARG-inference methods could be designed which
rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
and subsequently use a principled approach to perform ARG inference for the second phase.
This idea resembles the tried and tested approach of hybrid backwards-in-time simulations 
that combine Wright-Fisher dynamics in the very recent past and coalescent simulations 
for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
approach has enabled large scale simulations while avoiding the documented biases of the
coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. 
The same approach could be thus used for ARG-inference where one can rely on a heuristic 
inference method for the recent past, while using a model-based approach beyond that cutoff point.

Finally, acknowledging the uncertainty associated with ARG inference does not
necessarily require a full-blown MCMC sampler. Instead, recognising those parts of
the ARG for which we lack information for confident reconstruction and resolving
them stochastically by sampling from the coalescent rather than completely arbitrarily
would already be a major step forward.

% this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
We believe that quantifying uncertainty does not have to be associated with sampling
from the space of all compatible genealogies. Uncertainty could also
be quantified and passed on as metadata to the nodes in the data format storing the ARG.

[Note: a point to make here is that the idea of combining hueristic large-scale 
inference with model based inference is analogous to the combination of 
coalescent simulation and forwards time simulation. We can also draw lessons
from developers of different methods cooperating and using the tskit library
to enable this.]

>>>>

Our final example concerns [TODO motivate this example; why is 
this interesting/useful? Presumably we don't have to jointly estimate
 the ARG with demography, but can do so post-hoc? But, couldn't 
you do that anyway if you had a full ARG using HKY?]
% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, with a known time to recombination,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the left or
right hand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.
However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.
This approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution and context to a dataset of genetic variation.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently.

@@

The key contribution of this paper is to rederive the classical KYF
ARG likelihood function in a way that supports partial and imprecise
input, using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, acknowledging that we 
do not necessarily observe or estimate these events, we calculate 
the possible contributions of various type of event to the inheritance
relationships that are provided as input.
Although we can associate a likelihood with any ARG in \texttt{tskit}
format, there are some caveats. By treating each edge independently we can
naturally deal with the various degrees of completeness to which inheritance
is documented by the different ARG inference methods: e.g., polytomies, absence
of locally unary coalescence nodes, unpreserved identify of internal nodes
across local trees, etc.
However, many of these imprecisions violate the implicit
assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time and therefore recombination events.
Without these unary nodes, our algorithm will detect a recombination
event along both coalescing lineages in case of a (sub)tree height changes along
the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG research, but 
are estimated by \texttt{tsinfer}, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.

[A paragraph about possililties for inference]

[A paragraph about why what we've presented is useful, and a good starting
point for other research. We provide an efficient algorithm and 
reference implementation]

% This can be mitigated by limiting
% the number of `inferrable' recombination events per coalescence event to 1
% (see fig \ref{sup:fig:rec-correction}).
% \comment{I'm not editing this in since I'm not sure I've got it right:
% I think that as written, our algorithm would count an extra recombination
% and and extra coalescence for any edge that got split up by removing the unary bits.
% This could be corrected for in the algorithm, but it would require extra bookkeeping,
% and make the likelihood not so easily a product over edges. (?)}
% 
% \comment{this was previously part of the discussion, have grouped it here.}
% Although their presence is crucial to our ability to embed any local tree in the
% ARG, the concept and their importance are still relatively new (but see
% \citet{wong_general_2023}).
% We have briefly touched upon the observation that in
% the presence of unary nodes we seem to augment the left-to-right logic of the SMC
% with some additional information. Yet, we believe that the topic of locally
% unary nodes in general requires further exploration.
% And more specifically, we see it as future work to fully quantify to what
% extent the absence of these locally unary nodes will affect the shape of the
% likelihood surface given any set of parameters.

@@

% [JK: this needs significant work]
% Here, we have introduced a scalable algorithm to compute likelihoods
% under the SMC for a general class of ARGs. By formulating the SMC backwards in time
% we were able to associate a single likelihood value with each edge in the ARG.
% The compatibility of the algorithm with the \texttt{tskit} library allows us to
% exploit its associated efficiencies both for the current work and its future
% applications. This work will contribute significantly
% to ARG-based analysis becoming a standard part of the population
% and statistical genetics toolkit.
% In particular, we see two main applications for the work presented here.
% Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
% model and an associated set of parameters, will help bridge the gap between the tremendous recent
% advances in ARG inference and the current state of the art statistical inference of
% past demography.

% The second and probably the most important application of the ability to associate a more
% general class of ARGs to a principled model is the ability to boost the current efforts
% in ARG inference. 
% % great starting point for MCMC
% Any inferred ARG can be used a good starting point
% for a future MCMC-sampler. This potentially implies a massive runtime reduction by
% reducing the necessary burn-in period. Furthermore, by integrating out the exact time
% to the recombination event we further restrict the size of the ARG space that needs to
% be explored. This MCMC-sampler would again rely on the tree sequence data structure to
% efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
% % hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
% To further improve scalability, hybrid ARG-inference methods could be designed which
% rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
% and subsequently use a principled approach to perform ARG inference for the second phase.
% This idea resembles the tried and tested approach of hybrid backwards-in-time simulations 
% that combine Wright-Fisher dynamics in the very recent past and coalescent simulations 
% for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
% approach has enabled large scale simulations while avoiding the documented biases of the
% coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. 
% The same approach could be thus used for ARG-inference where one can rely on a heuristic 
% inference method for the recent past, while using a model-based approach beyond that cutoff point.

% Finally, acknowledging the uncertainty associated with ARG inference does not
% necessarily require a full-blown MCMC sampler. Instead, recognising those parts of
% the ARG for which we lack information for confident reconstruction and resolving
% them stochastically by sampling from the coalescent rather than completely arbitrarily
% would already be a major step forward.

% % this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
% We believe that quantifying uncertainty does not have to be associated with sampling
% from the space of all compatible genealogies. Uncertainty could also
% be quantified and passed on as metadata to the nodes in the data format storing the ARG.

% [Note: a point to make here is that the idea of combining hueristic large-scale 
% inference with model based inference is analogous to the combination of 
% coalescent simulation and forwards time simulation. We can also draw lessons
% from developers of different methods cooperating and using the tskit library
% to enable this.]
